---
title: "ggcomplex: Comprehensive Guide"
author: "Ruidong Yang"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_width: 10
    fig_height: 7
vignette: >
  %\VignetteIndexEntry{ggcomplex: Comprehensive Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse    = TRUE,
  comment     = "#>",
  fig.align   = "center",
  fig.width   = 10,
  fig.height  = 7,
  warning     = FALSE,
  message     = FALSE
)
```

# Introduction

**ggcomplex** is an R package for building complex composite figures ---
heatmaps with dendrograms, side annotations, bubble charts, and more --- using
a pipeline-style API built on top of *ggplot2*, *ggtree*, and *aplot*.

The core workflow follows an **Anchor + Satellites** pattern:

1. Create a **main plot** (`gg_heatmap`, `gg_dotplot`, or `gg_matrix`).
2. Optionally add **dendrograms** (`add_row_tree`, `add_col_tree`).
3. Attach **annotation tracks** (bar, tile, boxplot, violin, line, text, histogram).
4. **Print** or **save** the assembled composite figure.

```{r load}
library(ggcomplex)
library(ggplot2)
library(ggnewscale)   # optional: enables legends for circular ring annotations
```


# Simulated Data

We will use simulated gene expression data throughout this guide.

```{r data}
set.seed(2024)
n_genes   <- 20
n_samples <- 12

mat <- matrix(rnorm(n_genes * n_samples, mean = 0, sd = 2),
              nrow = n_genes,
              dimnames = list(paste0("Gene", 1:n_genes),
                              paste0("Sample", 1:n_samples)))

row_meta <- data.frame(
  gene   = rownames(mat),
  score  = runif(n_genes, 0, 10),
  group  = sample(c("Pathway_A", "Pathway_B", "Pathway_C"), n_genes, replace = TRUE),
  pvalue = runif(n_genes, 0.001, 0.05),
  stringsAsFactors = FALSE
)

col_meta <- data.frame(
  sample  = colnames(mat),
  batch   = rep(c("Batch1", "Batch2"), each = n_samples / 2),
  quality = rnorm(n_samples, 85, 5),
  stringsAsFactors = FALSE
)
```


# Data Pre-processing: `scale_matrix()`

`scale_matrix()` transforms a numeric matrix *before* plotting.
It supports **9 methods**:

| Method | Description |
|:---|:---|
| `"none"` | No transformation (default) |
| `"row_zscore"` | Row-wise Z-score standardisation |
| `"col_zscore"` | Column-wise Z-score |
| `"log2"` | log~2~(x + pseudocount) |
| `"log10"` | log~10~(x + pseudocount) |
| `"minmax"` | Row-wise min-max scaling to [0, 1] |
| `"col_minmax"` | Column-wise min-max scaling |
| `"center"` | Row-wise centering (subtract row mean) |
| `"col_center"` | Column-wise centering |

```{r scale-matrix}
mat_z <- scale_matrix(mat, method = "row_zscore")
cat("Z-score range:", round(range(mat_z), 2), "\n")

mat_mm <- scale_matrix(mat, method = "minmax")
cat("Min-max range:", round(range(mat_mm), 2), "\n")

count_mat <- matrix(rpois(n_genes * n_samples, lambda = 50),
                    nrow = n_genes,
                    dimnames = dimnames(mat))
mat_log <- scale_matrix(count_mat, method = "log2", pseudocount = 1)
cat("Log2 range:", round(range(mat_log), 2), "\n")
```


# Main Plot Types

## `gg_heatmap()` --- Tile Heatmap

The primary entry point. Every parameter is demonstrated below.

### Basic heatmap

```{r heatmap-basic}
obj <- gg_heatmap(mat_z)
obj
```

### Full parameter showcase

```{r heatmap-full, fig.height=8}
obj <- gg_heatmap(
  mat_z,
  fill_palette    = gc_palette("coolwarm"),
  fill_label      = "Z-score",
  fill_limits     = c(-3, 3),
  fill_midpoint   = 0,
  na_color        = "grey50",
  border_color    = "white",
  border_width    = 0.4,
  show_values     = TRUE,
  value_format    = "%.1f",
  value_size      = 2,
  value_color     = "grey20",
  show_row_names  = TRUE,
  show_col_names  = TRUE,
  row_names_size  = 7,
  col_names_size  = 7,
  col_names_angle = 60,
  row_names_side  = "right",
  col_names_side  = "bottom",
  row_names_face  = "italic",
  col_names_face  = "bold",
  legend_position = "right",
  title           = "Gene Expression Heatmap",
  base_size       = 10
)
obj
```

### Circle cell heatmap (correlation matrix)

`cell_shape = "circle"` draws circular markers instead of tiles --- ideal for
correlation matrices.

```{r heatmap-circle, fig.width=8, fig.height=8}
cor_mat <- cor(mat)

gg_heatmap(
  cor_mat,
  cell_shape      = "circle",
  fill_palette    = gc_palette("RdBu"),
  fill_label      = "Pearson r",
  fill_limits     = c(-1, 1),
  fill_midpoint   = 0,
  border_color    = NA,
  col_names_angle = 45,
  title           = "Sample Correlation (Circle Cells)"
)
```


## `gg_dotplot()` --- Bubble Chart

Encodes **two** numeric dimensions: colour and size.

### Basic dotplot

```{r dotplot-basic}
pmat <- matrix(runif(n_genes * n_samples, 0.001, 0.1),
               nrow = n_genes,
               dimnames = dimnames(mat))

gg_dotplot(mat_z, size_mat = pmat,
           color_label = "Z-score", size_label = "-log10(p)")
```

### Full parameter showcase

```{r dotplot-full, fig.height=8}
gg_dotplot(
  mat_z,
  size_mat        = -log10(pmat),
  color_palette   = gc_palette("PuOr"),
  color_label     = "Expression",
  size_label      = "Significance",
  size_range      = c(0.5, 6),
  size_threshold  = 1.0,
  size_reverse    = FALSE,
  point_shape     = 21,
  point_stroke    = 0.4,
  bg_color        = "grey97",
  bg_border       = "grey85",
  na_color        = "grey50",
  fill_limits     = c(-3, 3),
  fill_midpoint   = 0,
  show_values     = FALSE,
  show_row_names  = TRUE,
  show_col_names  = TRUE,
  row_names_size  = 7,
  col_names_size  = 7,
  col_names_angle = 45,
  row_names_side  = "left",
  col_names_side  = "bottom",
  legend_position = "right",
  title           = "Enrichment Dot Plot (shape 21 with fill)"
)
```


## `gg_matrix()` --- Custom Geom

The most flexible constructor. Pass any `geom_fn` and `mapping`.

### Using `geom_point` with colour scale auto-detection

```{r matrix-point}
gg_matrix(
  mat_z,
  geom_fn      = geom_point,
  mapping      = aes(color = .data$value, size = abs(.data$value)),
  fill_palette = gc_palette("spectral"),
  fill_label   = "Z-score",
  scale_type   = "auto",
  na_color     = "grey80",
  title        = "Custom Matrix (geom_point, auto color scale)"
)
```

### Adding extra layers

```{r matrix-extra}
gg_matrix(
  mat_z,
  geom_fn      = geom_tile,
  mapping      = aes(fill = .data$value),
  fill_palette = gc_palette("viridis"),
  fill_label   = "Z-score",
  extra_layers = list(
    geom_text(aes(label = ifelse(abs(.data$value) > 2, "*", "")),
              color = "white", size = 4)
  ),
  title        = "Matrix with Extra Annotation Layer"
)
```


# Dendrograms

## `add_row_tree()` / `add_col_tree()`

### Default clustering

```{r tree-default}
gg_heatmap(mat_z, fill_palette = gc_palette("RdBu")) %>%
  add_row_tree() %>%
  add_col_tree()
```

### Full parameter showcase

```{r tree-full, fig.height=9}
gg_heatmap(mat_z, fill_palette = gc_palette("RdYlBu"),
           fill_label = "Z-score") %>%
  add_row_tree(
    dist_method     = "correlation",
    hclust_method   = "ward.D2",
    width           = 0.2,
    line_size       = 0.6,
    color           = "grey40",
    linetype        = "solid",
    k               = 3,
    k_colors        = gc_cat_palette("npg", 3),
    show_tip_labels = FALSE,
    tip_label_size  = 2
  ) %>%
  add_col_tree(
    dist_method     = "euclidean",
    hclust_method   = "complete",
    height          = 0.12,
    line_size       = 0.5,
    linetype        = "solid",
    k               = 2,
    k_colors        = c("steelblue", "firebrick")
  )
```

### External pre-computed tree

```{r tree-external}
ext_hc <- hclust(as.dist(1 - cor(t(mat_z))), method = "average")

gg_heatmap(mat_z, fill_palette = gc_palette("PiYG")) %>%
  add_row_tree(tree = ext_hc, line_size = 0.8, color = "darkgreen")
```

### Custom distance function

```{r tree-custom-dist}
my_dist <- function(x) as.dist(1 - cor(t(x), method = "spearman"))

gg_heatmap(mat_z, fill_palette = gc_palette("BrBG")) %>%
  add_row_tree(dist_fn = my_dist, k = 4)
```


# Annotations

All annotation functions share a common parameter set:

| Parameter | Description |
|:---|:---|
| `palette` | Discrete colour vector for `scale_fill/color_manual` |
| `continuous_palette` | 2/3-colour vector for gradient fills |
| `fill` / `color` / `alpha` | Fixed aesthetic overrides |
| `border_color` / `border_width` | Border styling |
| `show_legend` | Show / hide this track's legend |
| `anno_title` | Small axis title for the annotation panel |
| `ylim` | Limit the numeric axis range |
| `extra_layers` | List of additional ggplot2 layers |

Prepare long-format data for boxplot / violin annotations:

```{r anno-longdata}
expr_long_row <- do.call(rbind, lapply(1:n_genes, function(i) {
  data.frame(gene = rownames(mat)[i], expr = mat[i, ])
}))

expr_long_col <- do.call(rbind, lapply(1:n_samples, function(j) {
  data.frame(sample = colnames(mat)[j], expr = mat[, j])
}))
```

## Bar chart: `add_row_bar()` / `add_col_bar()`

```{r anno-bar, fig.height=8, fig.width=10}
gg_heatmap(mat_z, fill_palette = gc_palette("RdBu")) %>%
  add_row_tree(dist_method = "correlation", k = 3) %>%
  add_col_tree(k = 2) %>%
  add_row_bar(
    row_meta,
    aes(x = gene, y = score, fill = group),
    palette    = gc_cat_palette("jco", 3),
    anno_title = "Score",
    ylim       = c(0, 12),
    width      = 0.3,
    alpha      = 0.85,
    side       = "right"
  ) %>%
  add_col_bar(
    col_meta,
    aes(x = sample, y = quality),
    fill       = "steelblue",
    anno_title = "Quality",
    ylim       = c(70, 100),
    height     = 0.18,
    side       = "top"
  )
```

## Tile strip: `add_row_tile()` / `add_col_tile()`

```{r anno-tile, fig.height=8, fig.width=10}
gg_heatmap(mat_z, fill_palette = gc_palette("RdBu")) %>%
  add_row_tree(k = 3) %>%
  add_col_tree() %>%
  add_row_tile(
    row_meta,
    aes(x = gene, y = 1, fill = group),
    palette      = gc_cat_palette("lancet", 3),
    border_color = "white",
    border_width = 0.8,
    anno_title   = "Pathway",
    width        = 0.06
  ) %>%
  add_col_tile(
    col_meta,
    aes(x = sample, y = 1, fill = batch),
    palette      = gc_cat_palette("nejm", 2),
    border_color = "white",
    border_width = 0.8,
    anno_title   = "Batch",
    height       = 0.04
  )
```

## Dot/Point: `add_row_dot()` / `add_col_dot()`

```{r anno-dot, fig.height=8, fig.width=10}
gg_heatmap(mat_z, fill_palette = gc_palette("RdBu")) %>%
  add_row_tree(k = 3) %>%
  add_col_tree() %>%
  add_row_dot(
    row_meta,
    aes(x = gene, y = score, color = group),
    palette    = gc_cat_palette("jco", 3),
    anno_title = "Score",
    width      = 0.25,
    side       = "right"
  ) %>%
  add_col_dot(
    col_meta,
    aes(x = sample, y = quality, color = batch),
    palette    = gc_cat_palette("nejm", 2),
    anno_title = "Quality",
    height     = 0.15,
    side       = "top"
  )
```

## Boxplot: `add_row_box()` / `add_col_box()`

```{r anno-box, fig.height=8, fig.width=10}
gg_heatmap(mat_z) %>%
  add_row_tree() %>%
  add_col_tree() %>%
  add_row_box(
    expr_long_row,
    aes(x = gene, y = expr),
    fill       = "lightyellow",
    color      = "grey50",
    anno_title = "Row Dist.",
    width      = 0.3,
    side       = "right"
  ) %>%
  add_col_box(
    expr_long_col,
    aes(x = sample, y = expr),
    fill       = "#D6EAF8",
    color      = "grey50",
    anno_title = "Col Dist.",
    height     = 0.2,
    side       = "top"
  )
```

## Violin: `add_row_violin()` / `add_col_violin()`

```{r anno-violin, fig.height=8, fig.width=10}
gg_heatmap(mat_z) %>%
  add_row_tree() %>%
  add_col_tree() %>%
  add_row_violin(
    expr_long_row,
    aes(x = gene, y = expr, fill = gene),
    show_legend = FALSE,
    alpha       = 0.6,
    anno_title  = "Row Violin",
    width       = 0.3,
    side        = "right"
  ) %>%
  add_col_violin(
    expr_long_col,
    aes(x = sample, y = expr, fill = sample),
    show_legend = FALSE,
    alpha       = 0.6,
    anno_title  = "Col Violin",
    height      = 0.2,
    side        = "top"
  )
```

## Line: `add_row_line()` / `add_col_line()`

```{r anno-line, fig.height=8, fig.width=10}
gg_heatmap(mat_z) %>%
  add_row_tree() %>%
  add_col_tree() %>%
  add_row_line(
    row_meta,
    aes(x = gene, y = score),
    color            = "darkgreen",
    line_show_points = TRUE,
    line_point_size  = 1.5,
    anno_title       = "Score Trend",
    width            = 0.25,
    side             = "right"
  ) %>%
  add_col_line(
    col_meta,
    aes(x = sample, y = quality),
    color            = "darkred",
    line_show_points = TRUE,
    line_point_size  = 2,
    anno_title       = "QC Score",
    height           = 0.15,
    side             = "top"
  )
```

## Text: `add_row_text()` / `add_col_text()`

```{r anno-text, fig.height=8, fig.width=10}
col_meta$label <- paste0("B", ifelse(col_meta$batch == "Batch1", 1, 2))

gg_heatmap(mat_z) %>%
  add_row_tree() %>%
  add_col_tree() %>%
  add_row_text(
    row_meta,
    aes(x = gene, y = 1, label = gene),
    text_size     = 2.5,
    text_fontface = "italic",
    text_angle    = 0,
    color         = "grey30",
    width         = 0.12,
    side          = "right"
  ) %>%
  add_col_text(
    col_meta,
    aes(x = sample, y = 1, label = label),
    text_size     = 3,
    text_fontface = "bold",
    text_angle    = 0,
    color         = "steelblue",
    height        = 0.05,
    side          = "top"
  )
```

## Histogram: `add_row_histogram()` / `add_col_histogram()`

```{r anno-hist, fig.height=8, fig.width=10}
gg_heatmap(mat_z) %>%
  add_row_tree() %>%
  add_col_tree() %>%
  add_row_histogram(
    row_meta,
    aes(x = gene, y = score),
    fill       = "coral",
    alpha      = 0.7,
    anno_title = "Score",
    ylim       = c(0, 12),
    width      = 0.2,
    side       = "right"
  ) %>%
  add_col_histogram(
    col_meta,
    aes(x = sample, y = quality),
    fill       = "#76D7C4",
    alpha      = 0.7,
    anno_title = "Quality",
    ylim       = c(70, 100),
    height     = 0.15,
    side       = "top"
  )
```

## Continuous palette annotation

When annotations map a continuous variable to fill, use `continuous_palette`
instead of the discrete `palette`.

```{r anno-continuous, fig.height=8, fig.width=10}
gg_heatmap(mat_z, fill_palette = gc_palette("RdBu")) %>%
  add_row_tree(k = 3) %>%
  add_row_bar(
    row_meta,
    aes(x = gene, y = pvalue, fill = pvalue),
    continuous_palette = c("white", "firebrick"),
    anno_title         = "P-value",
    width              = 0.18
  ) %>%
  add_col_dot(
    col_meta,
    aes(x = sample, y = quality, color = quality),
    continuous_palette = c("steelblue", "firebrick"),
    anno_title         = "Quality",
    height             = 0.15,
    side               = "top"
  )
```


# Panel Borders

Both the main heatmap and annotation panels support optional border frames
via `panel_border_color` and `panel_border_width`. Borders help visually
separate elements and give the figure a polished look.

## Main plot border

```{r border-main, fig.height=8}
gg_heatmap(mat_z,
           fill_palette       = gc_palette("RdBu"),
           panel_border_color = "grey30",
           panel_border_width = 1) %>%
  add_row_tree(k = 3) %>%
  add_col_tree(k = 2)
```

## Annotation panel borders

Each annotation track can independently enable a border:

```{r border-anno, fig.height=9}
gg_heatmap(mat_z,
           fill_palette       = gc_palette("RdYlBu"),
           fill_label         = "Z-score",
           panel_border_color = "grey40",
           panel_border_width = 0.8) %>%
  add_row_tree(k = 3) %>%
  add_col_tree(k = 2) %>%
  add_row_tile(
    row_meta, aes(x = gene, y = 1, fill = group),
    palette            = gc_cat_palette("jco", 3),
    anno_title         = "Pathway",
    panel_border_color = "grey40",
    panel_border_width = 0.8,
    width              = 0.06
  ) %>%
  add_row_bar(
    row_meta, aes(x = gene, y = score, fill = group),
    palette            = gc_cat_palette("jco", 3),
    anno_title         = "Score",
    ylim               = c(0, 12),
    panel_border_color = "grey40",
    panel_border_width = 0.8,
    width              = 0.25,
    show_legend        = FALSE
  ) %>%
  add_col_tile(
    col_meta, aes(x = sample, y = 1, fill = batch),
    palette            = gc_cat_palette("lancet", 2),
    anno_title         = "Batch",
    panel_border_color = "grey40",
    panel_border_width = 0.8,
    height             = 0.04
  ) %>%
  add_col_bar(
    col_meta, aes(x = sample, y = quality),
    fill               = "#4DBBD5",
    anno_title         = "Quality",
    ylim               = c(70, 100),
    panel_border_color = "grey40",
    panel_border_width = 0.8,
    height             = 0.15
  )
```


## Panel Spacing (gap between main plot and subplots)

Use `panel_spacing` in the main plot constructor to set a global default gap
(in pt) between the main heatmap and all surrounding panels. Each annotation or
tree function also accepts a `gap` argument that overrides the global default
for that individual panel.

```{r panel-spacing, fig.height=6, fig.width=10}
mat_sp <- matrix(rnorm(8 * 6), nrow = 8,
                 dimnames = list(paste0("G", 1:8), paste0("S", 1:6)))
row_df_sp <- data.frame(gene = paste0("G", 1:8),
                        score = runif(8, 2, 10))
col_df_sp <- data.frame(sample = paste0("S", 1:6),
                        quality = runif(6, 70, 100))

gg_heatmap(mat_sp, panel_spacing = 8) %>%
  add_row_tree() %>%
  add_col_tree() %>%
  add_row_bar(row_df_sp, aes(x = gene, y = score),
              fill = "steelblue", anno_title = "Score") %>%
  add_col_bar(col_df_sp, aes(x = sample, y = quality),
              fill = "#E64B35", anno_title = "Quality", ylim = c(70, 100))
```

Individual `gap` override per panel:

```{r panel-gap-override, fig.height=6, fig.width=10}
gg_heatmap(mat_sp, panel_spacing = 4) %>%
  add_row_tree(gap = 0) %>%
  add_col_tree(gap = 12) %>%
  add_row_bar(row_df_sp, aes(x = gene, y = score),
              fill = "steelblue", anno_title = "Score", gap = 10) %>%
  add_col_bar(col_df_sp, aes(x = sample, y = quality),
              fill = "#E64B35", anno_title = "Quality",
              ylim = c(70, 100), gap = 0)
```


# Colour Palettes

## Continuous palettes: `gc_palette()`

17 built-in palettes for heatmap gradients.

```{r palette-continuous}
gc_palette_names()
```

```{r palette-demo-cont, fig.height=3, fig.width=9}
demo <- expand.grid(x = 1:20, palette = c("RdBu", "viridis", "plasma", "YlOrRd"))
demo$value <- rep(seq(-3, 3, length.out = 20), 4)

ggplot(demo, aes(x = x, y = palette, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B") +
  theme_minimal() +
  labs(title = "Palette Preview (conceptual)")
```

## Categorical palettes: `gc_cat_palette()`

26 palettes covering RColorBrewer, ggsci journal, genomics, and sci-fi themes.

```{r palette-cat, fig.height=5, fig.width=8}
pal_names <- c("npg", "jco", "lancet", "nejm", "d3", "set1")
plot_list <- lapply(pal_names, function(nm) {
  cols <- gc_cat_palette(nm)
  df <- data.frame(x = seq_along(cols), color = cols)
  ggplot(df, aes(x = x, y = 1, fill = color)) +
    geom_tile(show.legend = FALSE) +
    scale_fill_identity() +
    labs(title = nm) +
    theme_void() +
    theme(plot.title = element_text(hjust = 0.5, size = 9))
})
gridExtra::grid.arrange(grobs = plot_list, ncol = 2)
```

Interpolation for more colours than the palette contains:

```{r palette-interpolate}
cat("NPG (10 colours):", gc_cat_palette("npg"), "\n")
cat("NPG (15 interpolated):", gc_cat_palette("npg", n = 15), "\n")
```


# Utility Functions

## `set_title()` --- Global titles

```{r set-title, fig.height=7}
gg_heatmap(mat_z) %>%
  add_row_tree() %>%
  add_col_tree() %>%
  set_title(
    title    = "My Composite Figure",
    subtitle = "20 genes x 12 samples (Z-score normalised)",
    caption  = "Generated by ggcomplex"
  )
```

## `summary()` --- Object inspection

```{r summary}
obj <- gg_heatmap(mat_z) %>%
  add_row_tree(k = 3) %>%
  add_col_tree() %>%
  add_row_bar(row_meta, aes(x = gene, y = score, fill = group),
              palette = gc_cat_palette("npg", 3)) %>%
  add_col_tile(col_meta, aes(x = sample, y = 1, fill = batch),
               palette = gc_cat_palette("lancet", 2))

summary(obj)
```

## `assemble_plot()` --- Manual assembly

`assemble_plot()` returns the raw `aplot` object for further manipulation:

```{r assemble}
ap <- assemble_plot(obj)
class(ap)
```

## `subset_rows()` / `subset_cols()` --- Subsetting

```{r subset, fig.height=5}
obj_small <- gg_heatmap(mat_z) %>%
  add_row_tree() %>%
  add_col_tree() %>%
  subset_rows(paste0("Gene", 1:10)) %>%
  subset_cols(paste0("Sample", 1:6))
obj_small
```

## `gc_save()` --- Saving to file

```{r gc-save, eval=FALSE}
gc_save(obj, "composite_heatmap.pdf", width = 14, height = 10, dpi = 300)
gc_save(obj, "composite_heatmap.png", width = 14, height = 10, dpi = 300)
```


# Circular Heatmaps

ggcomplex supports **circular (polar) heatmaps** where columns are arranged
around the circle (angular axis) and rows form concentric rings (radial axis).
This layout is especially useful for large sample panels or when presenting
data in a visually compact way.

> **Tip:** Install the optional
> [`ggnewscale`](https://CRAN.R-project.org/package=ggnewscale) package to
> enable automatic legend generation for annotation rings (`add_outer_ring()`
> with `palette` or `continuous_palette`). Without it, ring colours are still
> drawn but legends are suppressed.
>
> ```r
> install.packages("ggnewscale")
> ```

## `gg_circular()` --- Basic Circular Heatmap

```{r circ-basic, fig.width=8, fig.height=8}
set.seed(2024)
circ_mat <- matrix(rnorm(8 * 24), 8, 24,
                   dimnames = list(paste0("Gene", 1:8),
                                   paste0("S", 1:24)))

gg_circular(circ_mat,
            fill_palette  = gc_palette("RdBu"),
            fill_label    = "Expression",
            fill_midpoint = 0,
            inner_radius  = 3,
            track_height  = 1,
            title         = "Basic Circular Heatmap (full circle)")
```

### Key parameters

| Parameter | Description |
|:---|:---|
| `inner_radius` | Radius of the inner hole (larger = bigger centre) |
| `track_height` | Radial height of each row ring |
| `ring_gap` | Default radial gap between heatmap and annotation rings |
| `gap_degree` | Size of the opening in degrees (0 = full, 30 = 30° gap) |
| `start_angle` | Starting angle in radians (`NULL` = auto for gap) |
| `direction` | `1` = clockwise, `-1` = counter-clockwise |
| `show_row_labels` | Display row names in the gap opening |
| `show_col_labels` | Display column names around the outer edge |

### Opening / gap

Set `gap_degree` to create an arc-shaped opening at the bottom of the
circle, commonly used in publication-quality circular heatmaps:

```{r circ-gap, fig.width=8, fig.height=8}
gg_circular(circ_mat,
            fill_palette    = gc_palette("RdBu"),
            gap_degree      = 40,
            inner_radius    = 4,
            track_height    = 0.8,
            show_row_labels = TRUE,
            row_label_size  = 2.5,
            title           = "Circular Heatmap with 40-degree Gap")
```

## `add_circular_tree()` --- Circular Dendrogram

`add_circular_tree(which = "col")` clusters columns and draws a fan-shaped
dendrogram **inside** the inner circle. `which = "row"` reorders rows only
(no visual tree).

```{r circ-tree, fig.width=8, fig.height=8}
gg_circular(circ_mat,
            fill_palette = c("navy", "white", "firebrick"),
            inner_radius = 4,
            track_height = 0.8) %>%
  add_circular_tree(which = "col",
                    dist_method   = "euclidean",
                    hclust_method = "ward.D2",
                    color         = "grey40",
                    line_size     = 0.3)
```

Row clustering (reorder only, no inner tree):

```{r circ-row-cluster, fig.width=8, fig.height=8}
gg_circular(circ_mat,
            fill_palette   = gc_palette("PuOr"),
            inner_radius   = 4,
            show_row_labels = TRUE,
            row_label_size  = 2.5) %>%
  add_circular_tree(which = "row",
                    dist_method = "correlation") %>%
  add_circular_tree(which = "col", color = "brown")
```

## `add_outer_ring()` --- Annotation Rings

Annotation rings are added concentrically **outside** the heatmap. Four geom
types are supported: `"tile"`, `"bar"`, `"point"`, and `"text"`.

When `ggnewscale` is installed, tile and histogram rings with `palette` or
`continuous_palette` will **automatically generate legends** in the plot,
making it easy to identify what each colour ring represents. Column/row labels
and sector names are rendered **horizontally** for better readability,
regardless of their angular position. Dynamic label offsets prevent overlap
between outer ring annotations and labels.

### Tile ring (discrete groups)

```{r circ-ring-tile, fig.width=9, fig.height=9}
circ_col_meta <- data.frame(
  sample = colnames(circ_mat),
  group  = rep(c("Control", "Treatment"), each = 12),
  score  = runif(24, 0, 100),
  stringsAsFactors = FALSE
)

gg_circular(circ_mat,
            fill_palette = gc_palette("RdBu"),
            inner_radius = 3,
            track_height = 0.8) %>%
  add_circular_tree(which = "col", color = "grey50") %>%
  add_outer_ring(circ_col_meta,
                 col_var  = "sample",
                 fill_var = "group",
                 geom     = "tile",
                 palette  = c(Control = "#4DAF4A", Treatment = "#E41A1C"),
                 ring_label = "Group",
                 height   = 0.8)
```

### Tile ring (continuous palette)

```{r circ-ring-cont, fig.width=9, fig.height=9}
circ_pval_meta <- data.frame(
  sample = colnames(circ_mat),
  pvalue = runif(24, 0, 1)
)

gg_circular(circ_mat,
            fill_palette = gc_palette("RdBu"),
            inner_radius = 3) %>%
  add_outer_ring(circ_pval_meta,
                 col_var  = "sample",
                 fill_var = "pvalue",
                 geom     = "tile",
                 continuous_palette = c("white", "darkgreen"),
                 ring_label = "-log10(p)",
                 height   = 0.8)
```

### Bar ring (numeric values)

```{r circ-ring-bar, fig.width=9, fig.height=9}
gg_circular(circ_mat,
            fill_palette = gc_palette("RdBu"),
            inner_radius = 3,
            track_height = 0.8) %>%
  add_circular_tree(which = "col") %>%
  add_outer_ring(circ_col_meta,
                 col_var   = "sample",
                 value_var = "score",
                 geom      = "bar",
                 fill      = "steelblue",
                 ring_label = "Score",
                 height    = 1.5)
```

### Text ring

```{r circ-ring-text, fig.width=9, fig.height=9}
circ_label_meta <- data.frame(
  sample = colnames(circ_mat),
  label  = paste0("Lab", 1:24)
)

gg_circular(circ_mat,
            inner_radius = 3,
            track_height = 0.8) %>%
  add_outer_ring(circ_label_meta,
                 col_var   = "sample",
                 label_var = "label",
                 geom      = "text",
                 height    = 1)
```

## Circular Heatmap Borders

`heatmap_border_color` and `heatmap_border_width` draw thick arc and radial
borders around the heatmap ring, giving a polished, publication-ready look.

```{r circ-border, fig.width=8, fig.height=8}
gg_circular(circ_mat,
            fill_palette         = gc_palette("PuOr"),
            inner_radius         = 3,
            gap_degree           = 30,
            heatmap_border_color = "black",
            heatmap_border_width = 1.2,
            title                = "Circular Heatmap with Borders") %>%
  add_circular_tree(which = "col", color = "grey40")
```


## Column Split (Sectors)

`col_split` divides columns into named groups shown as separate angular
sectors, separated by `split_gap` x-units. An inner ring automatically
displays group labels **horizontally** (always readable, regardless of
angular position). Combined with `heatmap_border_color` this produces
the classic split circular heatmap look.

| Parameter | Description |
|:---|:---|
| `col_split` | Named character/factor vector (one per column, names = colnames) |
| `split_gap` | Angular gap between sectors (x-units, default 1) |
| `heatmap_border_color` | Colour for sector borders (`NA` = no border) |
| `heatmap_border_width` | Line width of the sector border |

### Basic split

```{r circ-split-basic, fig.width=9, fig.height=9}
circ_groups <- setNames(
  rep(c("Group A", "Group B", "Group C"), c(8, 8, 8)),
  colnames(circ_mat)
)

gg_circular(circ_mat,
            fill_palette         = c("#E66101", "white", "#1B7837"),
            fill_midpoint        = 0,
            gap_degree           = 50,
            inner_radius         = 3,
            track_height         = 0.8,
            col_split            = circ_groups,
            split_gap            = 1.5,
            heatmap_border_color = "grey20",
            heatmap_border_width = 0.9,
            show_col_labels      = TRUE,
            col_label_size       = 2.2,
            title                = "Three-Sector Circular Heatmap")
```

### Split + dendrogram + annotation rings

```{r circ-split-full, fig.width=10, fig.height=10}
circ_groups4 <- setNames(
  rep(c("A", "B", "C", "D"), each = 6),
  colnames(circ_mat)
)

gg_circular(
    circ_mat,
    fill_palette         = gc_palette("RdBu"),
    fill_label           = "Z-score",
    fill_midpoint        = 0,
    inner_radius         = 4,
    track_height         = 0.7,
    gap_degree           = 40,
    col_split            = circ_groups4,
    split_gap            = 1.2,
    heatmap_border_color = "grey30",
    heatmap_border_width = 0.8,
    show_row_labels      = TRUE,
    row_label_size       = 2,
    title                = "Split Circular Heatmap (4 sectors)"
  ) %>%
  add_circular_tree(which = "col",
                    hclust_method = "ward.D2",
                    color = "grey40") %>%
  add_outer_ring(circ_col_meta,
                 col_var  = "sample",
                 fill_var = "group",
                 geom     = "tile",
                 palette  = c(Control = "#4DAF4A", Treatment = "#E41A1C"),
                 ring_label = "Group",
                 height   = 0.6) %>%
  add_outer_ring(circ_col_meta,
                 col_var   = "sample",
                 value_var = "score",
                 geom      = "bar",
                 fill      = "#4DBBD5",
                 ring_label = "Score",
                 height    = 1.2)
```

### Full-circle split (no gap)

```{r circ-split-full-circle, fig.width=9, fig.height=9}
gg_circular(circ_mat,
            col_split            = circ_groups4,
            split_gap            = 1,
            heatmap_border_color = "grey30",
            title                = "Full-Circle with Four Sectors")
```


## Inner Links (Chord Diagram)

`add_inner_links()` draws curved Bézier connections inside the inner circle,
creating a chord-diagram / network-graph effect. Each link connects two column
positions, curving through the centre.

| Parameter | Description |
|:---|:---|
| `links` | Data frame with *from* and *to* columns (column names) |
| `color_var` | Optional column to map to link colour |
| `continuous_palette` | Gradient colours for a continuous `color_var` |
| `palette` | Named colours for a discrete `color_var` |
| `linewidth_var` | Optional column to map to line width |
| `curvature` | 0 = flat, 1 = dips to the very centre (default 0.7) |
| `alpha` | Transparency (default 0.5) |

### Correlation links (continuous colour + width)

```{r circ-links-basic, fig.width=9, fig.height=9}
circ_cor <- cor(circ_mat)
cor_pairs <- which(abs(circ_cor) > 0.3 & upper.tri(circ_cor), arr.ind = TRUE)
cor_links <- data.frame(
  from = colnames(circ_mat)[cor_pairs[, 1]],
  to   = colnames(circ_mat)[cor_pairs[, 2]],
  cor  = circ_cor[cor_pairs]
)

gg_circular(circ_mat,
            fill_palette = gc_palette("RdBu"),
            inner_radius = 4,
            gap_degree   = 30,
            title        = "Correlation Links (|r| > 0.3)") %>%
  add_inner_links(cor_links,
                  from_var  = "from", to_var = "to",
                  color_var = "cor",
                  continuous_palette = c("steelblue", "grey90", "firebrick"),
                  linewidth_var     = "cor",
                  linewidth_range   = c(0.3, 1.5),
                  alpha = 0.6, curvature = 0.7)
```

### Discrete-colour links

```{r circ-links-discrete, fig.width=9, fig.height=9}
net_links <- data.frame(
  from = c("S1", "S3", "S5", "S8", "S10", "S14", "S18", "S22"),
  to   = c("S12", "S17", "S20", "S16", "S24", "S2",  "S6",  "S11"),
  type = rep(c("activation", "inhibition"), each = 4)
)

gg_circular(circ_mat,
            fill_palette = gc_palette("PuOr"),
            inner_radius = 4,
            gap_degree   = 30,
            title        = "Network-style Links (discrete)") %>%
  add_inner_links(net_links,
                  from_var  = "from", to_var = "to",
                  color_var = "type",
                  palette   = c(activation = "#E41A1C", inhibition = "#377EB8"),
                  linewidth = 0.8, alpha = 0.65, curvature = 0.75)
```

### Links + tree + annotation ring

```{r circ-links-tree, fig.width=10, fig.height=10}
gg_circular(circ_mat,
            fill_palette         = gc_palette("RdBu"),
            inner_radius         = 5,
            gap_degree           = 40,
            heatmap_border_color = "grey30",
            heatmap_border_width = 0.8,
            title                = "Links + Tree + Ring") %>%
  add_circular_tree(which = "col", color = "grey50") %>%
  add_outer_ring(circ_col_meta,
                 col_var  = "sample", fill_var = "group", geom = "tile",
                 palette  = c(Control = "#4DAF4A", Treatment = "#E41A1C"),
                 ring_label = "Group", height = 0.6) %>%
  add_inner_links(cor_links,
                  from_var  = "from", to_var = "to",
                  color_var = "cor",
                  continuous_palette = c("blue", "grey80", "red"),
                  alpha = 0.4, curvature = 0.6)
```

### Links inside split sectors

```{r circ-links-split, fig.width=10, fig.height=10}
circ_groups4 <- setNames(
  rep(c("A", "B", "C", "D"), each = 6),
  colnames(circ_mat)
)

gg_circular(circ_mat,
            fill_palette         = c("#E66101", "white", "#1B7837"),
            inner_radius         = 4,
            gap_degree           = 50,
            col_split            = circ_groups4,
            split_gap            = 1.2,
            heatmap_border_color = "grey20",
            title                = "Split Sectors + Links") %>%
  add_inner_links(cor_links,
                  from_var  = "from", to_var = "to",
                  color_var = "cor",
                  continuous_palette = c("purple", "grey90", "orange"),
                  linewidth_var     = "cor",
                  linewidth_range   = c(0.3, 1.2),
                  alpha = 0.55, curvature = 0.8)
```


## Multi-ring Circular Example

A production-quality figure combining **all** circular features:
column split (4 sectors), gap opening, row & column dendrograms, inner
correlation links, heatmap borders, row labels, and **all 6 outer ring
types** (tile, bar, point, line, histogram, text).

With `ggnewscale` installed, this example will display legends for the
discrete tile ring ("Group") and the continuous tile ring ("P-value"),
as well as the histogram ring ("Mean Expr").

```{r circ-comprehensive-data}
circ_label_meta <- data.frame(
  sample = colnames(circ_mat),
  label  = paste0("Lab", 1:24)
)

circ_expr_meta <- data.frame(
  sample   = colnames(circ_mat),
  expr_avg = colMeans(circ_mat),
  fc       = rnorm(24, mean = 0.5, sd = 1.2)
)

circ_signal_meta <- data.frame(
  sample = colnames(circ_mat),
  signal = sin(seq(0, 4 * pi, length.out = 24)) * 2 + runif(24, -0.3, 0.3)
)
```

```{r circ-comprehensive, fig.width=12, fig.height=12}
gg_circular(
    circ_mat,
    fill_palette         = gc_palette("RdBu"),
    fill_label           = "Z-score",
    fill_midpoint        = 0,
    inner_radius         = 5,
    track_height         = 0.6,
    gap_degree           = 40,
    col_split            = circ_groups4,
    split_gap            = 1.2,
    heatmap_border_color = "grey20",
    heatmap_border_width = 0.8,
    show_row_labels      = TRUE,
    row_label_size       = 2.2,
    ring_gap             = 0.3,
    title                = "All Features: 4 sectors + 6 ring types + links + trees + borders"
  ) %>%
  # --- inner dendrograms ---
  add_circular_tree(which = "col",
                    hclust_method = "ward.D2",
                    color = "grey40") %>%
  add_circular_tree(which = "row",
                    dist_method = "correlation") %>%
  # --- inner correlation links (chord diagram) ---
  add_inner_links(cor_links,
                  from_var  = "from", to_var = "to",
                  color_var = "cor",
                  continuous_palette = c("steelblue", "grey80", "firebrick"),
                  alpha = 0.3, curvature = 0.7) %>%
  # --- Ring 1: discrete tile (group) ---
  add_outer_ring(circ_col_meta,
                 col_var    = "sample",
                 fill_var   = "group",
                 geom       = "tile",
                 palette    = c(Control = "#4DAF4A", Treatment = "#E41A1C"),
                 ring_label = "Group",
                 height     = 0.5) %>%
  # --- Ring 2: bar (score) ---
  add_outer_ring(circ_col_meta,
                 col_var    = "sample",
                 value_var  = "score",
                 geom       = "bar",
                 fill       = "#4DBBD5",
                 ring_label = "Score",
                 height     = 1.0) %>%
  # --- Ring 3: continuous tile (p-value heatmap) ---
  add_outer_ring(circ_pval_meta,
                 col_var            = "sample",
                 fill_var           = "pvalue",
                 geom               = "tile",
                 continuous_palette = c("lightyellow", "darkred"),
                 ring_label         = "P-value",
                 height             = 0.5) %>%
  # --- Ring 4: point/dot (fold change) ---
  add_outer_ring(circ_expr_meta,
                 col_var    = "sample",
                 value_var  = "fc",
                 geom       = "point",
                 fill       = "#7570B3",
                 ring_label = "FC",
                 height     = 0.8) %>%
  # --- Ring 5: line (signal trend) ---
  add_outer_ring(circ_signal_meta,
                 col_var    = "sample",
                 value_var  = "signal",
                 geom       = "line",
                 fill       = "#D95F02",
                 ring_label = "Signal",
                 height     = 1.0) %>%
  # --- Ring 6: histogram (expression mean, coloured by group) ---
  add_outer_ring(circ_expr_meta,
                 col_var            = "sample",
                 value_var          = "expr_avg",
                 fill_var           = "expr_avg",
                 geom               = "histogram",
                 continuous_palette = c("#F7FCB1", "#31A354"),
                 ring_label         = "Mean Expr",
                 height             = 1.0) %>%
  # --- Ring 7: text labels (outermost) ---
  add_outer_ring(circ_label_meta,
                 col_var   = "sample",
                 label_var = "label",
                 geom      = "text",
                 height    = 1.0)
```

### `summary()` for circular objects

```{r circ-summary}
circ_obj <- gg_circular(circ_mat, inner_radius = 3) %>%
  add_circular_tree(which = "col") %>%
  add_outer_ring(circ_col_meta,
                 col_var = "sample", fill_var = "group",
                 geom = "tile",
                 palette = c(Control = "#4DAF4A", Treatment = "#E41A1C")) %>%
  add_outer_ring(circ_col_meta,
                 col_var = "sample", value_var = "score",
                 geom = "bar", fill = "steelblue")

summary(circ_obj)
```


# Comprehensive Example (Linear)

Putting everything together in a single production-quality figure that uses
**all 8 annotation types** (bar, tile, dot, boxplot, violin, line, text,
histogram) distributed across **all 4 sides** with panel borders on every
element.

| Side | Annotations |
|:---|:---|
| **Left** | tile (pathway), bar (score), violin (distribution), dot (p-value) |
| **Right** | boxplot (expression), histogram (score), text (labels) |
| **Top** | tile (batch), bar (quality), line (QC trend), dot (quality) |
| **Bottom** | boxplot (expression), violin (distribution), histogram (QC), text (labels) |

```{r comprehensive-data}
expr_long_row <- do.call(rbind, lapply(1:n_genes, function(i) {
  data.frame(gene = rownames(mat)[i], expr = mat[i, ])
}))
expr_long_col <- do.call(rbind, lapply(1:n_samples, function(j) {
  data.frame(sample = colnames(mat)[j], expr = mat[, j])
}))
```

```{r comprehensive, fig.width=18, fig.height=14}
final <- gg_heatmap(
    scale_matrix(mat, "row_zscore"),
    fill_palette       = gc_palette("RdBu"),
    fill_label         = "Z-score",
    fill_limits        = c(-3, 3),
    fill_midpoint      = 0,
    border_color       = "white",
    border_width       = 0.3,
    row_names_side     = "right",
    row_names_face     = "italic",
    row_names_size     = 7,
    col_names_angle    = 50,
    base_size          = 9,
    panel_border_color = "grey30",
    panel_border_width = 0.8
  ) %>%
  # --- dendrograms -------------------------------------------------------
  add_row_tree(
    dist_method   = "correlation",
    hclust_method = "ward.D2",
    k             = 3,
    k_colors      = gc_cat_palette("npg", 3),
    line_size     = 0.6,
    width         = 0.18
  ) %>%
  add_col_tree(
    dist_method   = "euclidean",
    hclust_method = "complete",
    k             = 2,
    k_colors      = c("#0073C2", "#EFC000"),
    height        = 0.10
  ) %>%
  # --- LEFT: tile + bar + violin (side = "left") -------------------------
  add_row_tile(
    row_meta, aes(x = gene, y = 1, fill = group),
    palette            = gc_cat_palette("jco", 3),
    anno_title         = "Pathway",
    panel_border_color = "grey50",
    panel_border_width = 0.5,
    width              = 0.05,
    side               = "left"
  ) %>%
  add_row_bar(
    row_meta, aes(x = gene, y = score, fill = group),
    palette            = gc_cat_palette("jco", 3),
    anno_title         = "Score",
    ylim               = c(0, 12),
    panel_border_color = "grey50",
    panel_border_width = 0.5,
    width              = 0.22,
    show_legend        = FALSE,
    side               = "left"
  ) %>%
  add_row_violin(
    expr_long_row, aes(x = gene, y = expr, fill = gene),
    show_legend        = FALSE,
    alpha              = 0.5,
    anno_title         = "Distr.",
    panel_border_color = "grey50",
    panel_border_width = 0.5,
    width              = 0.22,
    side               = "left"
  ) %>%
  add_row_dot(
    row_meta, aes(x = gene, y = pvalue, color = pvalue),
    continuous_palette = c("steelblue", "firebrick"),
    anno_title         = "P-val",
    panel_border_color = "grey50",
    panel_border_width = 0.5,
    width              = 0.15,
    side               = "left"
  ) %>%
  # --- RIGHT: boxplot + histogram + text (side = "right") ----------------
  add_row_box(
    expr_long_row, aes(x = gene, y = expr),
    fill               = "lightyellow",
    color              = "grey50",
    anno_title         = "Boxplot",
    panel_border_color = "grey50",
    panel_border_width = 0.5,
    width              = 0.22,
    side               = "right"
  ) %>%
  add_row_histogram(
    row_meta, aes(x = gene, y = score),
    fill               = "coral",
    alpha              = 0.7,
    anno_title         = "Score",
    ylim               = c(0, 12),
    panel_border_color = "grey50",
    panel_border_width = 0.5,
    width              = 0.15,
    side               = "right"
  ) %>%
  add_row_text(
    row_meta, aes(x = gene, y = 1, label = gene),
    text_size          = 2,
    text_fontface      = "italic",
    color              = "grey30",
    panel_border_color = "grey50",
    panel_border_width = 0.5,
    width              = 0.10,
    side               = "right"
  ) %>%
  # --- TOP: tile + bar + line (side = "top") -----------------------------
  add_col_tile(
    col_meta, aes(x = sample, y = 1, fill = batch),
    palette            = gc_cat_palette("lancet", 2),
    anno_title         = "Batch",
    panel_border_color = "grey50",
    panel_border_width = 0.5,
    height             = 0.03,
    side               = "top"
  ) %>%
  add_col_bar(
    col_meta, aes(x = sample, y = quality),
    fill               = "#4DBBD5",
    anno_title         = "Quality",
    ylim               = c(70, 100),
    panel_border_color = "grey50",
    panel_border_width = 0.5,
    height             = 0.12,
    side               = "top"
  ) %>%
  add_col_line(
    col_meta, aes(x = sample, y = quality),
    color              = "darkred",
    line_show_points   = TRUE,
    line_point_size    = 1.5,
    anno_title         = "QC Trend",
    panel_border_color = "grey50",
    panel_border_width = 0.5,
    height             = 0.12,
    side               = "top"
  ) %>%
  add_col_dot(
    col_meta, aes(x = sample, y = quality, color = quality),
    continuous_palette = c("green4", "orange"),
    anno_title         = "QC Dot",
    panel_border_color = "grey50",
    panel_border_width = 0.5,
    height             = 0.10,
    side               = "top"
  ) %>%
  # --- BOTTOM: boxplot + violin + histogram + text (side = "bottom") -----
  add_col_box(
    expr_long_col, aes(x = sample, y = expr),
    fill               = "lightyellow",
    color              = "grey50",
    anno_title         = "Expression",
    panel_border_color = "grey50",
    panel_border_width = 0.5,
    height             = 0.15,
    side               = "bottom"
  ) %>%
  add_col_violin(
    expr_long_col, aes(x = sample, y = expr, fill = sample),
    show_legend        = FALSE,
    alpha              = 0.5,
    anno_title         = "Violin",
    panel_border_color = "grey50",
    panel_border_width = 0.5,
    height             = 0.15,
    side               = "bottom"
  ) %>%
  add_col_histogram(
    col_meta, aes(x = sample, y = quality),
    fill               = "#00A087",
    alpha              = 0.7,
    anno_title         = "QC Hist",
    ylim               = c(70, 100),
    panel_border_color = "grey50",
    panel_border_width = 0.5,
    height             = 0.12,
    side               = "bottom"
  ) %>%
  add_col_text(
    col_meta, aes(x = sample, y = 1, label = sample),
    text_size          = 2.5,
    text_fontface      = "bold",
    color              = "grey30",
    panel_border_color = "grey50",
    panel_border_width = 0.5,
    height             = 0.04,
    side               = "bottom"
  ) %>%
  set_title(
    title    = "Full-Feature Composite Heatmap",
    subtitle = "All 8 annotation types across all 4 sides: bar, tile, dot, violin, boxplot, line, histogram, text"
  )

final
```


# Alternative: Unified Interface `geom_add()`

All the composite plots above can also be built with a single function —
`geom_add()`. It wraps every `add_row_*()` / `add_col_*()` helper into one
entry-point. Use whichever style you prefer; the result is identical.

| Parameter | Description |
|:---|:---|
| `geom` | `"bar"`, `"tile"`, `"dot"`, `"boxplot"`, `"violin"`, `"line"`, `"text"`, `"histogram"`, or `"tree"` |
| `side` | `"left"`, `"right"`, `"top"`, or `"bottom"` |
| `data` | Annotation data frame (not needed for `"tree"`) |
| `mapping` | `aes()` mapping (not needed for `"tree"`) |
| `size` | Panel width (left/right) or height (top/bottom) |
| `title` | Axis title for the panel |
| `...` | All other params forwarded (e.g. `k`, `dist_method`, `text_size`, etc.) |

## Basic usage

```{r geom-add-basic, fig.height=8, fig.width=10}
gg_heatmap(mat_z, fill_palette = gc_palette("RdBu")) %>%
  geom_add("tree",  side = "left", k = 3, k_colors = gc_cat_palette("npg", 3)) %>%
  geom_add("tree",  side = "top") %>%
  geom_add("bar",   side = "right", data = row_meta,
           mapping = aes(x = gene, y = score, fill = group),
           palette = gc_cat_palette("jco", 3), title = "Score",
           ylim = c(0, 12), size = 0.3) %>%
  geom_add("tile",  side = "left", data = row_meta,
           mapping = aes(x = gene, y = 1, fill = group),
           palette = gc_cat_palette("lancet", 3), title = "Pathway",
           size = 0.05) %>%
  geom_add("dot",   side = "top", data = col_meta,
           mapping = aes(x = sample, y = quality, color = quality),
           continuous_palette = c("steelblue", "firebrick"),
           title = "Quality", size = 0.12)
```

## Full composite with `geom_add()`

The comprehensive example from the previous section, rewritten entirely
with `geom_add()`:

```{r geom-add-full, fig.width=14, fig.height=10}
gg_heatmap(mat_z, fill_palette = gc_palette("RdBu"),
           panel_border_color = "grey30") %>%
  geom_add("tree", side = "left",
           dist_method = "correlation", hclust_method = "ward.D2",
           k = 3, k_colors = gc_cat_palette("npg", 3)) %>%
  geom_add("tree", side = "top", k = 2) %>%
  # --- left side ---
  geom_add("tile", side = "left", data = row_meta,
           mapping = aes(x = gene, y = 1, fill = group),
           palette = gc_cat_palette("jco", 3), title = "Pathway",
           size = 0.05) %>%
  geom_add("violin", side = "left", data = expr_long_row,
           mapping = aes(x = gene, y = expr, fill = gene),
           show_legend = FALSE, alpha = 0.5, title = "Distr.",
           size = 0.2) %>%
  # --- right side ---
  geom_add("bar", side = "right", data = row_meta,
           mapping = aes(x = gene, y = score),
           fill = "steelblue", title = "Score",
           ylim = c(0, 12), size = 0.25) %>%
  geom_add("dot", side = "right", data = row_meta,
           mapping = aes(x = gene, y = pvalue, color = pvalue),
           continuous_palette = c("grey80", "red3"),
           title = "P-val", size = 0.15) %>%
  geom_add("text", side = "right", data = row_meta,
           mapping = aes(x = gene, y = 1, label = gene),
           color = "grey30", size = 0.10,
           text_size = 2, text_fontface = "italic") %>%
  # --- top side ---
  geom_add("tile", side = "top", data = col_meta,
           mapping = aes(x = sample, y = 1, fill = batch),
           palette = gc_cat_palette("lancet", 2), title = "Batch",
           size = 0.03) %>%
  geom_add("bar", side = "top", data = col_meta,
           mapping = aes(x = sample, y = quality),
           fill = "#4DBBD5", title = "Quality",
           ylim = c(70, 100), size = 0.12) %>%
  # --- bottom side ---
  geom_add("boxplot", side = "bottom", data = expr_long_col,
           mapping = aes(x = sample, y = expr),
           fill = "lightyellow", color = "grey50",
           title = "Expr.", size = 0.15) %>%
  geom_add("line", side = "bottom", data = col_meta,
           mapping = aes(x = sample, y = quality),
           color = "darkred", title = "QC Trend",
           line_show_points = TRUE, size = 0.12) %>%
  geom_add("histogram", side = "bottom", data = col_meta,
           mapping = aes(x = sample, y = quality),
           fill = "#00A087", title = "QC Hist",
           ylim = c(70, 100), size = 0.10)
```


# Session Info

```{r session}
sessionInfo()
```
